Lattice Boltzmann simulations of magnetically responsive\\ ellipsoidal particle stabilized emulsions

### 2025-03-21 chapters/analysis_techniques.tex
Changes 1. Updated aim 1 with first person wording 2. Updated aim 2 with first person wording and modified conclusion 3. Updated aim 3 with full size conclusion
Changes 1. Addressed some image sizes and phrasing consistency
Draft 1 complete

In this work, the structure factor derived domain size, geometric pore size distribution and tortuosity will be 
utilized to characterize the microstructure. The nematic order parameter will be used to characterize the particle 
order to the magnetic field. The average angle to the interface will be utilized to characterize how the particles 
ordering to the field affect the interface position. Finally, the radial distribution function will be used to probe 
how differences in ordering change the packing of particles on the interface. The time evolution of viscosity will 
provide insight into the rheological properties of the bijel, while the aforementioned analysis tools in addition to 
proportion of particles on the interface will be calculated to provide insight into how the bijel reacts to shear, 
parallel and perpendicular to the magnetic field. A more detailed description of the implementation of each method is 
shown in subsequent sections.

@@

Particles are the cause of bijels but they can also make characterization of domain size, curvature, pore size 
distribution and tortuosity more difficult as particles may change these parameters. An extrapolation scheme is 
required to connect the interface one different sides of the particles and to fill in the particles themselves so 
as to create a more accurate interface for other techniques.

@@

We use a method based on that outlined in Chan and Thornton. We create our distance function by performing an Eularian Distance Transform on 
our order parameter data $\phi$ to obtain $\phi_{edt}$. \cite{chan_channel_2012} We smooth $\phi_{edt}$ using a boxcar algorithm with side length 3. 
We then normalize the distance from the interface in $\phi_{edt}$ using the characteristic length scale of the system defined as $\Sigma = \frac{V}{A_{int}}$
to obtain $\bar{\phi}_{edt}$ and a normalized distance from the interface, $\bar{r}$. A schematic of this workflow is demonstrated in Figure \ref{fig:csd_prep_viz}.

\begin{figure}
    \centering
    \includegraphics[scale = 0.4]{figures/analysis/csd_prep.png}
    \caption{Schematics detailing the order of operations of the steps to calculate the channel size distribution.}
    \label{fig:csd_prep_viz}
\end{figure}

>>>>

Many techniques used in the bijel literature such as various definitions of the domain size,
nematic order parameter, radial distribution function and shear stress calculation were used in the characterization
of bijels in this work. \cite{kendon_inertial_2001, jansen_bijels_2011, gunther_timescales_2014} Other techniques such
as the curvature and channel size distribution have also been used to characterize the microstructure of bijels. 
\cite{reeves_quantitative_2016,chan_channel_2012} Additional techniques such as the interfacial angle
and the Steinhardt order parameters. \cite{karthikeyan_formation_2024,steinhardt_bond-orientational_1983, mickel_shortcomings_2013}
This section introduces each method and they are calculated. The implementation of each function can be found in my 
\href{https://github.com/nkarthi95/schiller_lab_tools}{github repository}.

@@

Particles are the cause of bijels but they can also make characterization of domain size, curvature, pore size 
distribution and tortuosity more difficult as the presence of particles affects measurement of these parameters. 
An extrapolation scheme is 
required to connect the interface one different sides of the particles and to fill in the particles themselves so 
as to create a more accurate interface for other techniques.

@@

We use a method based on that outlined in Chan and Thornton. We create our distance function by performing an Eulerian Distance Transform on 
our order parameter data $\phi$ to obtain $\phi_{edt}$. \cite{chan_channel_2012} We smooth $\phi_{edt}$ using a boxcar algorithm with side length 3. 
We then normalize the distance from the interface in $\phi_{edt}$ using the characteristic length scale of the system defined as $\Sigma = \frac{V}{A_{int}}$
to obtain $\bar{\phi}_{edt}$ and a normalized distance from the interface, $\bar{r}$. A schematic of this workflow is demonstrated in Figure \ref{fig:csd_prep_viz}.

\begin{figure}
    \centering
    \includegraphics[scale = 0.5]{figures/analysis/csd_prep.png}
    \caption{Schematics detailing the order of operations of the steps to calculate the channel size distribution.}
    \label{fig:csd_prep_viz}
\end{figure}

### 2025-03-21 chapters/final_remarks.tex
Changes 1. Addressed some image sizes and phrasing consistency
Draft 1 complete

This work utilized 2 ellipsoidal particle geometries chosen for comparisons to previous literature using this 
particle geometry. Particles based on cellulose nanocrystals or graphene nanoplates are now in use to fabricate
particle stabilized emulsions.These particles have also been shown to have capillary bridging and particle stacking,
not seen in the particles used in this work. These particles in bulk have been shown to have some intrinsic ordering
that can be predicted using onsager theory. An investigation into how using rods or plate like particles would be 
instructive in identifying if onsager theory can be used in bijel formation and its link to bijel microstructure. 
\cite{tan_2d_2021}

@@

Another avenue of exploration would be the use of gradient or rotating magnetic fields in place of the constant magnetic 
fields used here. One study on bijel microstructure showed that a gradient in the particle volume fraction can be used to 
create a gradient in the eventual domain size. A gradient field may be able to generate a gradient in the nematic order 
parameter, affecting the particle packing of the bijel at different heights and varying the pore size as a function of 
height in the field gradient direction. In a ferrofluid or solution of magnetic colloids, a rotating field can assemble 
particles into chains or rings. In the context of bijel structural response, this can be used to tune bijel microstructure 
as this process can control the unjamming and rejamming of the particle monolayer, allowing for greater control over the 
resulting bijel microstructure than a constant field would have. In these simulations, the frequency of rotation would 
likely need to be tuned based upon how quickly particles respond to field in the bijel.

@@

>>>>

This work utilized 2 ellipsoidal particle geometries chosen for comparisons to previous literature using this 
particle geometry. Particles based on cellulose nanocrystals or graphene nanoplates are now in use to fabricate
particle stabilized emulsions. These particles have also been shown to have capillary bridging and particle stacking,
not seen in the particles used in this work. These particles in bulk have been shown to have some intrinsic ordering
that can be predicted using onsager theory. An investigation into how using rods or plate like particles would be 
instructive in identifying if onsager theory can be used in bijel formation and its link to bijel microstructure. 
\cite{tan_2d_2021}

@@

Another avenue of exploration would be the use of gradient or rotating magnetic fields in place of the constant magnetic 
fields used here. One study on bijel microstructure showed that a gradient in the particle volume fraction can be used to 
create a gradient in the eventual domain size. A gradient field may be able to generate nematic order 
parameter gradients, affecting the particle packing of the bijel at different heights and varying the pore size as a function of 
height in the field gradient direction. In a ferrofluid or solution of magnetic colloids, a rotating field can assemble 
particles into chains or rings. In the context of bijel structural response, this can be used to tune bijel microstructure 
as this process can control the unjamming and rejamming of the particle monolayer, allowing for greater control over the 
resulting bijel microstructure than a constant field would have. In these simulations, the frequency of rotation would 
likely need to be tuned based upon how quickly particles respond to field in the bijel.

@@

This research was supported in part through the use of DARWIN computing system: DARWIN - A Resource for Computational 
and Data-intensive Research at the University of Delaware and in the Delaware Region, Rudolf Eigenmann, Benjamin E. 
Bagozzi, Arthi Jayaraman, William Totten, and Cathy H. Wu, University of Delaware, 2021, URL: https://udspace.udel.edu/handle/19716/29071

### 2025-03-21 chapters/lit_review.tex
Changes 1. Addressed some image sizes and phrasing consistency
Draft 1 complete

\begin{figure}
    \centering
    \includegraphics[scale = 0.2]{figures/literature_review/state_diagram.jpg}
    \caption{Formation criteria for a particle stabilized emulsion that can form bijel. \cite{velankar_non-equilibrium_2015}
            Reproduced from Velankar Soft Matter, 2015,11, 8393-8403 using license number 1551968-1.}
    \label{fig:state_diagram_particle_emulsions}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale = 2]{figures/literature_review/bijel_under_shear.jpeg}
    \caption{Schematic of a bijel with hard-sphere particles undergoing shear, demonstrating migration and detachment of 
             particles at the interface. \cite{bonaccorso_shear_2020} Reproduced from Bonaccorso et al. AIP Advances 1 
             September 2020; 10 (9): 095304, under the Creative Commons CC BY License.}
    \label{fig:bijel_under_shear}
\end{figure}

>>>>

\begin{figure}
    \centering
    \includegraphics[scale = 0.3]{figures/literature_review/state_diagram.jpg}
    \caption{Formation criteria for a particle stabilized emulsion that can form bijel. \cite{velankar_non-equilibrium_2015}
            Reproduced from Velankar Soft Matter, 2015,11, 8393-8403 using license number 1551968-1.}
    \label{fig:state_diagram_particle_emulsions}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[scale = 3]{figures/literature_review/bijel_under_shear.jpeg}
    \caption{Schematic of a bijel with hard-sphere particles undergoing shear, demonstrating migration and detachment of 
             particles at the interface. \cite{bonaccorso_shear_2020} Reproduced from Bonaccorso et al. AIP Advances 1 
             September 2020; 10 (9): 095304, under the Creative Commons CC BY License.}
    \label{fig:bijel_under_shear}
\end{figure}

### 2025-03-21 chapters/paper1.tex
Changes 1. Addressed some image sizes and phrasing consistency
Draft 1 complete

In LB simulations of spinodal decomposition, a characteristic length
scale can be obtained from the structure factor, defined in Section \ref{section:domain_size}. \cite{kendon_3d_1999,kendon_inertial_2001} 
It is worth noting that this definition of average domain size does not necessarily yield values that coincide with the typical
domain size observed visually in snapshots such as those shown in Fig.~\ref{fig:microstructure_viz}. However, in a dynamic
scaling regime, different measures of domain size are expected to differ only by a constant ratio. I found that the commonly 
used measure \(L_1(t)\) tends to overestimate the observed domain size. Numerically, the calculation of the spherically 
averaged structure factor is subject to poor statistics in the low \(k\)-shells, where the average \(|\vec{k}|\) of the 
lattice sites differs from the nominal shell radius. Additionally, the measurement of \(L(t)\) in simulations with
periodic boundary conditions is subject to finite size effects. More importantly, in suspensions of anisotropic particles, 
I cannot expect that the structure factor is isotropic. Therefore, I also used the 3D structure factor to calculate a lateral domain 
size in each Cartesian direction from the second moments also defined in Section \ref{section:domain_size}. 
\cite{jansen_bijels_2011,gunther_timescales_2014} I add the additional processing calculating a parallel and perpendicular domain size
from the results in each cartesian direction, defining $L_{\parallel}=L_z$ and $L_{\perp} = \frac{L_x+L_y}{2}$ I plot the structure factor of the structures in
Figure \ref{fig:structure_factor}.

@@

At first glance, the time-dependence of the domain size in Fig.
\ref{fig:domain_size} suggests that the domain size is not strongly
affected by the external magnetic field. In Fig.\ref{fig:D2a_B}a) I
plotted the domain size measure \(L_d\) normalized by \(R\), the larger
of the two radii of the ellipsoidal particles. The data points represent
the domain size measured at the final time step and averaged over three
independent simulation runs with the same parameters. The normalization
with the particle radius makes it apparent that the ellipsoidal
particles reduce the domain size of bijels. This observation is in line
with observations by Günther et al.~\cite{gunther_timescales_2014} and
indicates that -- relative to their volume -- ellipsoidal particles
stabilize a larger interface area than spherical particles. This is due
to the larger cross-sectional area of ellipsoidal particles at the same
volume. Furthermore, steric constraints prevent anisotropic particles to
pack less densely in the interface than spherical particles, as I will
discuss in more detail below.

@@

To measure the particle alignment quantitatively, I computed the
nematic order tensor using the method detailed in Section \ref{section:nematic_order_parameter} 
Figure \ref{fig:nematic_time} shows the time evolution of the nematic
order parameter \(S\) for the three different particle aspect ratios
\(\alpha\) and varying magnetic field strength \(\bar{B}\). The onset of
orientational order is nearly instantaneous and increases with
increasing field strength. For magnetic flux densities
\(\bar{B}\ge0.5\), the nematic order parameter saturates at
\(S\approx 1.0\) towards the end of the simulation. For the prolate
particles (\(\alpha=2.0\)), I observe a shoulder at around
\(10^4\Delta t\) and an intermittent plateau. The occurrence of this
plateau coincides with the slowed decay of the coarsening speed in Fig.
\ref{fig:coarsening_velocity}. This change in the orientational ordering
is even more distinct for the oblate particles, where the nematic order
decays intermittently before it increases again. The intermittent decay
for oblate particles, and the plateau for prolate particles, correlate
with the delayed onset of jamming in the direction perpendicular and
parallel to the magnetic field, respectively. These observations can be
interpreted as follows: The magnetic field aligns the particles with the
direction of the magnetic field early during the simulation. Due to this
alignment, the steric constraints between particles adsorbed at the
interface are reduced which allows further coarsening of the interfacial
area. However, during coarsening, the interfaces may re-orient
themselves and cause a capillary torque on the particles that competes
with the magnetic torque. This competition intermittently perturbs the
nematic order of the particles leading to the observed decay and plateau
of the nematic order parameter. Accordingly, the decay of \(S\) is most
pronounced at the weakest magnetic field \(\bar{B}=0.2\). To further corroborate 
this mechanism, I plot the time evolution of the orientational order of the interfaces.

@@

The radial distribution function is shown at time steps
\(t=1000\Delta t,\ 10000\Delta t,\ 100000\Delta t\). For ellipsoidal
particles, the characteristic distances between particles in contact are
\(2R_{\parallel}\), \(2R_{\perp}\), and \(R_{\parallel} + R_{\perp}\).
These distances are marked by dashed lines in Fig. \ref{fig:rdf}. We
observe that the largest peak of the RDF coincides with the distance
\(2R_\perp\) for both oblate and spherical particles, indicating that
oblate particles tend to be in contact at their circumference while
prolate particles preferentially align side-by-side. These arrangements
lead to closer packing of particles in the interface. The peaks are less
pronounced in the early stage of the simulations, where the particles
are more randomly oriented. Notably, there is a dip in the RDF of
prolate particles at \(2R_\parallel\) indicating that initially the
tip-to-tip configuration occurs more rarely. The growth of the peak
height over time shows that the magnetic fields promote this alignment
compared to the case without magnetic field. For oblate particles, a
shoulder develops around \(R_\parallel+R_\perp\), indicating that some
particles tilt out of the interface to reduce the covered interfacial
area. For prolate particles, the dip at \(2R_\parallel\) disappears over
time, indicating that more particles align tip-to-tip. This is
consistent with alignment of prolate particles with respect to the
magnetic field, which facilitates closer packing along both axes of the
particles. It further suggests that particles arrange in layers, which can also be observed in the snapshot in Fig.
\ref{fig:packing_viz}. Together with the analysis of nematic order
above, these results substantiate the proposed mechanisms that influence
the formation of bijels with anisotropic particles in magnetic fields:
The magnetic field aligns the particle dipole axis in the field
direction, and the particle re-orientation couples to interface
alignment due to the capillary interactions. The alignment of particles
reduces the steric constraints within the interface, which allows the
interface to shrink further and facilitates domain coarsening. In terms
of the time-evolution, the rotation of particles due to the magnetic
field occurs during the initial stages of the simulation, followed by
the local re-alignment of interfaces due to capillary interactions. In
the case of prolate particles, I observed the shrinking interface can
force particles to tilt out of their preferential alignment to facilitate
closer packing. Generally, our results show that when bijels stabilized
by anisotropic magnetic particles form under the influence of a magnetic
field, the domain size and tortuosity become anisotropic. Additionally,
I observe that the particle packing in the interface becomes more
ordered due to the alignment effects.

>>>>

In LB simulations of spinodal decomposition, a characteristic length
scale can be obtained from the structure factor, defined in Section \ref{section:domain_size}. \cite{kendon_3d_1999,kendon_inertial_2001} 
It is worth noting that this definition of average domain size does not necessarily yield values that coincide with the typical
domain size observed visually in snapshots such as those shown in Figure~\ref{fig:microstructure_viz}. However, in a dynamic
scaling regime, different measures of domain size are expected to differ only by a constant ratio. I found that the commonly 
used measure \(L_1(t)\) tends to overestimate the observed domain size. Numerically, the calculation of the spherically 
averaged structure factor is subject to poor statistics in the low \(k\)-shells, where the average \(|\vec{k}|\) of the 
lattice sites differs from the nominal shell radius. Additionally, the measurement of \(L(t)\) in simulations with
periodic boundary conditions is subject to finite size effects. More importantly, in suspensions of anisotropic particles, 
I cannot expect that the structure factor is isotropic. Therefore, I also used the 3D structure factor to calculate a lateral domain 
size in each Cartesian direction from the second moments also defined in Section \ref{section:domain_size}. 
\cite{jansen_bijels_2011,gunther_timescales_2014} I add the additional processing calculating a parallel and perpendicular domain size
from the results in each cartesian direction, defining $L_{\parallel}=L_z$ and $L_{\perp} = \frac{L_x+L_y}{2}$ I plot the structure factor of the structures in
Figure \ref{fig:structure_factor}.

@@

At first glance, the time-dependence of the domain size in Figure 
\ref{fig:domain_size} suggests that the domain size is not strongly
affected by the external magnetic field. In Figure \ref{fig:D2a_B}a) I
plotted the domain size measure \(L_d\) normalized by \(R\), the larger
of the two radii of the ellipsoidal particles. The data points represent
the domain size measured at the final time step and averaged over three
independent simulation runs with the same parameters. The normalization
with the particle radius makes it apparent that the ellipsoidal
particles reduce the domain size of bijels. This observation is in line
with observations by Günther et al.~\cite{gunther_timescales_2014} and
indicates that -- relative to their volume -- ellipsoidal particles
stabilize a larger interface area than spherical particles. This is due
to the larger cross-sectional area of ellipsoidal particles at the same
volume. Furthermore, steric constraints prevent anisotropic particles to
pack less densely in the interface than spherical particles, as I will
discuss in more detail below.

@@

To measure the particle alignment quantitatively, I computed the
nematic order tensor using the method detailed in Section \ref{section:nematic_order_parameter} 
Figure \ref{fig:nematic_time} shows the time evolution of the nematic
order parameter \(S\) for the three different particle aspect ratios
\(\alpha\) and varying magnetic field strength \(\bar{B}\). The onset of
orientational order is nearly instantaneous and increases with
increasing field strength. For magnetic flux densities
\(\bar{B}\ge0.5\), the nematic order parameter saturates at
\(S\approx 1.0\) towards the end of the simulation. For the prolate
particles (\(\alpha=2.0\)), I observe a shoulder at around
\(10^4\Delta t\) and an intermittent plateau. The occurrence of this
plateau coincides with the slowed decay of the coarsening speed in Figure
\ref{fig:coarsening_velocity}. This change in the orientational ordering
is even more distinct for the oblate particles, where the nematic order
decays intermittently before it increases again. The intermittent decay
for oblate particles, and the plateau for prolate particles, correlate
with the delayed onset of jamming in the direction perpendicular and
parallel to the magnetic field, respectively. These observations can be
interpreted as follows: The magnetic field aligns the particles with the
direction of the magnetic field early during the simulation. Due to this
alignment, the steric constraints between particles adsorbed at the
interface are reduced which allows further coarsening of the interfacial
area. However, during coarsening, the interfaces may re-orient
themselves and cause a capillary torque on the particles that competes
with the magnetic torque. This competition intermittently perturbs the
nematic order of the particles leading to the observed decay and plateau
of the nematic order parameter. Accordingly, the decay of \(S\) is most
pronounced at the weakest magnetic field \(\bar{B}=0.2\). To further corroborate 
this mechanism, I plot the time evolution of the orientational order of the interfaces.

@@

The radial distribution function is shown at time steps
\(t=1000\Delta t,\ 10000\Delta t,\ 100000\Delta t\). For ellipsoidal
particles, the characteristic distances between particles in contact are
\(2R_{\parallel}\), \(2R_{\perp}\), and \(R_{\parallel} + R_{\perp}\).
These distances are marked by dashed lines in Figure \ref{fig:rdf}. We
observe that the largest peak of the RDF coincides with the distance
\(2R_\perp\) for both oblate and spherical particles, indicating that
oblate particles tend to be in contact at their circumference while
prolate particles preferentially align side-by-side. These arrangements
lead to closer packing of particles in the interface. The peaks are less
pronounced in the early stage of the simulations, where the particles
are more randomly oriented. Notably, there is a dip in the RDF of
prolate particles at \(2R_\parallel\) indicating that initially the
tip-to-tip configuration occurs more rarely. The growth of the peak
height over time shows that the magnetic fields promote this alignment
compared to the case without magnetic field. For oblate particles, a
shoulder develops around \(R_\parallel+R_\perp\), indicating that some
particles tilt out of the interface to reduce the covered interfacial
area. For prolate particles, the dip at \(2R_\parallel\) disappears over
time, indicating that more particles align tip-to-tip. This is
consistent with alignment of prolate particles with respect to the
magnetic field, which facilitates closer packing along both axes of the
particles. It further suggests that particles arrange in layers, which can also be observed in the snapshot in Figure
\ref{fig:packing_viz}. Together with the analysis of nematic order
above, these results substantiate the proposed mechanisms that influence
the formation of bijels with anisotropic particles in magnetic fields:
The magnetic field aligns the particle dipole axis in the field
direction, and the particle re-orientation couples to interface
alignment due to the capillary interactions. The alignment of particles
reduces the steric constraints within the interface, which allows the
interface to shrink further and facilitates domain coarsening. In terms
of the time-evolution, the rotation of particles due to the magnetic
field occurs during the initial stages of the simulation, followed by
the local re-alignment of interfaces due to capillary interactions. In
the case of prolate particles, I observed the shrinking interface can
force particles to tilt out of their preferential alignment to facilitate
closer packing. Generally, our results show that when bijels stabilized
by anisotropic magnetic particles form under the influence of a magnetic
field, the domain size and tortuosity become anisotropic. Additionally,
I observe that the particle packing in the interface becomes more
ordered due to the alignment effects.
